# GTEx preprocessing

Load libraries
```{r}
options(download.file.method="wget")

if (!requireNamespace("PLIER", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    install.packages("devtools")
  }
  library(devtools)
  install_github("wgmao/PLIER")
}

library(bigstatsr)
library(data.table)
library(dplyr)
library(rsvd)
library(glmnet)
library(Matrix)
library(knitr)
library(here)
library(PLIER)
```

```{r load_config}
source(here("R/solvers.R"))
source(here("R/utilsNew.R"))
source(here("config.R"))
```

Output
```{r}
output_data_dir <- config$GTEx$DATASET_FOLDER
dir.create(output_data_dir, showWarnings = FALSE, recursive = TRUE)

output_pathways_dir <- file.path(config$GENERAL$DATA_DIR, "pathways")
dir.create(output_pathways_dir, showWarnings = FALSE, recursive = TRUE)
```

Download GTEx 
```{r download-gtex, message=FALSE, warning=FALSE}
url <- config$GTEx$URL
dest_dir <-  config$GTEx$DATASET_FOLDER
dest_gz  <- file.path(dest_dir, basename(url))

if (!file.exists(dest_gz)) {
  dir.create(dest_dir, recursive = TRUE, showWarnings = FALSE)
  download.file(url, dest_gz, mode = "wb")
  message("Downloaded to: ", dest_gz)
} else {
  message("File already exists, skipping download.")
}
```

Download prior-knowledge
```{r}
gmtList=list()

gmt_dir <- output_pathways_dir

if (!file.exists(file.path(gmt_dir, "WikiPathway_2021_Human.gmt"))) {
  download.file(
    "https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=WikiPathway_2021_Human",
    file.path(gmt_dir, "WikiPathway_2021_Human.gmt"),
    method = "wget", quiet = TRUE
  )
}
if (!file.exists(file.path(gmt_dir, "MGI_Mammalian_Phenotype_Level_4_2021.gmt"))) {
  download.file(
    "https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=MGI_Mammalian_Phenotype_Level_4_2021",
    file.path(gmt_dir, "MGI_Mammalian_Phenotype_Level_4_2021.gmt"),
    method = "wget", quiet = TRUE
  )
}
if (!file.exists(file.path(gmt_dir, "GO_Biological_Process_2023.gmt"))) {
  download.file(
    "https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=GO_Biological_Process_2023",
    file.path(gmt_dir, "GO_Biological_Process_2023.gmt"),
    method = "wget", quiet = TRUE
  )
}
if (!file.exists(file.path(gmt_dir, "Reactome_2022.gmt"))) {
  download.file(
    "https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=Reactome_2022",
    file.path(gmt_dir, "Reactome_2022.gmt"),
    method = "wget", quiet = TRUE
  )
}
if (!file.exists(file.path(gmt_dir, "CellMarker_2024.gmt"))) {
  download.file(
    "https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=CellMarker_2024",
    file.path(gmt_dir, "CellMarker_2024.gmt"),
    method = "wget", quiet = TRUE
  )
}

pathMat = gmtListToSparseMat(gmtList[c(5)])
```

Process GTEx
```{r process-gtex, message=FALSE, warning=FALSE}
exprs_path  <- file.path(config$GTEx$DATASET_FOLDER, 'GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct.gz')
output_file <- config$GTEx$DATASET_FILE

if (!file.exists(output_file)) {
  dir.create(dirname(output_file), recursive = TRUE, showWarnings = FALSE)
  
  exprs_data <- read.table(exprs_path, header = TRUE, sep = "\t", skip = 2, check.names = FALSE)
  
  saveRDS(exprs_data, config$GTEx$DATASET_FILE)
  message("File successfully written to: ", config$GTEx$DATASET_FILE)
} else {
  message("Output file already exists. Skipping.")
}
```

Read in gtex RDS file 
```{r}
gtex=readRDS(here(config$GTEx$DATASET_FILE))
```

Aggregate in-place by 'description'
```{r}
gtex <- as.data.table(gtex)
aggregated_gtex <- gtex[, lapply(.SD, sum), by = Description, .SDcols = is.numeric]
```

Create a meta data file to attach to the FBM matrix
```{r}
meta <- list()
meta$rowNames <- aggregated_gtex$Description
meta$colNames <- colnames(aggregated_gtex)[-1]
data_mat <- as.matrix(aggregated_gtex[, -1])
```

Create the FBM
```{r}
fbm_file <- file.path(output_data_dir, "FBMgtex")
gtexFBM <- FBM(nrow = nrow(data_mat), ncol = ncol(data_mat), backingfile = fbm_file, create_bk = T)
```

Populate it with data
```{r}
block_size <- config$GENERAL$CHUNK_SIZE

n_blocks <- ceiling(nrow(aggregated_gtex) / block_size)

for (i in 1:n_blocks) {
  start_row <- (i-1) * block_size + 1
  end_row <- min(i * block_size, nrow(data_mat))
  
  gtexFBM[start_row:end_row, ] <- as.matrix(data_mat[start_row:end_row, ])
}
```

```{r}
cleanFBM(gtexFBM)
```

```{r}
rowStats=computeRowStatsFBM(gtexFBM)
plot(rowStats$row_means, rowStats$row_variances, log="y")
```

Filter the data. Doing aggressive filtering to reduce the number of genes.
```{r}
filterResult=filterFBM(gtexFBM, rowStats, mean_cutoff = config$GTEx$GENES_MEAN_CUTOFF, var_cutoff = config$GTEx$GENES_VAR_CUTOFF, backingfile =paste0(fbm_file, "_filtered")) 
gtexFBMfiltered=filterResult$fbm_filtered
meta$rowNames=meta$rowNames[filterResult$kept_rows]
rowStatsAll=rowStats
rowStats$row_means=rowStats$row_means[filterResult$kept_rows]
rowStats$row_variances=rowStats$row_variances[filterResult$kept_rows]
```

Do the z-scoring
```{r}
zscoreFBM(gtexFBMfiltered, rowStats = rowStats)
```

save the FBM object for quick loading
```{r}
gtexFBMfiltered$save()
saveRDS(meta, paste0(fbm_file, "_meta.RDS"))
```

K for SVD
```{r k, echo=TRUE}
n_genes   <- length(meta$rowNames)
n_samples <- length(meta$colNames)

SVD_K <- min(n_genes, n_samples) - 1
message(SVD_K)
```

Precompute the SVD
```{r}
# SVD_K as 17215 is taking forever
SVD_K = 200
gtexFBM.svd=big_randomSVD(gtexFBMfiltered, k = SVD_K)
save(gtexFBM.svd, file = file.path(output_data_dir, "svd.rds"))
```

Infer K from SVD using elbow method
```{r infer_k, echo=TRUE}
svd_list <- list(d=gtexFBM.svd$d)
PLIER_K <- num.pc(svd_list) * 2

message(paste0("K inferred for PLIER: ", PLIER_K))
```

Run simpleDecomp as the base model 
```{r}
decomp_file <- file.path(output_data_dir, "gtexSimpleDecompResult.RDS")

if (!file.exists(decomp_file)) {
  gtex.sdres <- simpleDecomp(gtexFBMfiltered, k = PLIER_K, svdres = gtexFBM.svd)
  saveRDS(gtex.sdres, file = decomp_file)
} else {
  gtex.sdres <- readRDS(decomp_file)
}
```

Create a pathMat object and save
```{r}
pathMat = gmtListToSparseMat(gmtList[c(5)])
```

Get a matched pathway set
The FBM version doesn't do any subsetting, everything has to match exactly
```{r}
matchedPaths=getMatchedPathwayMat(pathMat, meta$rowNames)
Chat=getChat(matchedPaths, method = "fast")
```

# Run PLIER2

Run PLIER2, setting `max.iter` low for demonstration
```{r}
gtex.plier2 <-  PLIERv2(
    gtexFBMfiltered,
    matchedPaths,
    sdres = gtex.sdres,
    Chat = Chat,
    doCrossval = TRUE,
    multiplier = config$GTEx$PLIER_PARAMS$MULTIPLIER,
    max.U.updates = config$GTEx$PLIER_PARAMS$MAX_U_UPDATES,
    max.iter = config$GTEx$PLIER_PARAMS$MAX_ITER,
    k = PLIER_K
    )

saveRDS(gtex.plier2, file = file.path(output_data_dir, "gtex_PLIER2.rds"))
```

# Run PLIER

Run PLIER with the same inputs than PLIER2

```{r}
library(PLIER)

gtex.plier = PLIER::PLIER(
    gtexFBMfiltered[], 
    as.matrix(matchedPaths), 
    svdres = gtexFBM.svd, 
    Chat = as.matrix(Chat), 
    doCrossval = TRUE, 
    max.iter = config$GTEx$PLIER_PARAMS$MAX_ITER,
    k = PLIER_K
  )

saveRDS(gtex.plier, file = file.path(output_data_dir, "gtex_PLIER.rds"))
```

