# GTEx preprocessing

Load libraries
```{r}
options(download.file.method="wget")

if (!requireNamespace("PLIER", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    install.packages("devtools")
  }
  library(devtools)
  install_github("wgmao/PLIER")
}

library(bigstatsr)
library(data.table)
library(dplyr)
library(rsvd)
library(glmnet)
library(Matrix)
library(knitr)
library(here)
library(PLIER)
```

Output 

```{r}
output_nb_path <- here("output/vignettes/plier2_testing/gtex")
dir.create(output_nb_path, showWarnings = FALSE, recursive = TRUE)
```

Load the local files
```{r}
source(here("R/solvers.R"))
source(here("R/utilsNew.R"))
```

Download GTEx 

```{r download-gtex, message=FALSE, warning=FALSE}
url      <- "https://storage.googleapis.com/adult-gtex/bulk-gex/v8/rna-seq/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct.gz"
dest_dir <- "data"
dest_gz  <- file.path(dest_dir, basename(url))

if (!file.exists(dest_gz)) {
  dir.create(dest_dir, recursive = TRUE, showWarnings = FALSE)
  download.file(url, dest_gz, mode = "wb")
  message("Downloaded to: ", dest_gz)
} else {
  message("File already exists, skipping download.")
}
```

Process GTEx

```{r process-gtex, message=FALSE, warning=FALSE}
exprs_path  <- "data/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct.gz"
output_file <- "data/bulk-gex_v8_rna-seq_GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.rds"

if (!file.exists(output_file)) {
  dir.create(dirname(output_file), recursive = TRUE, showWarnings = FALSE)
  
  exprs_data <- read.table(exprs_path, header = TRUE, sep = "\t", skip = 2, check.names = FALSE) %>%
    dplyr::rename(
      gene_ens_id = Name,
      gene_symbol = Description
    )
  
  saveRDS(exprs_data, output_file)
  message("File successfully written to: ", output_file)
} else {
  message("Output file already exists. Skipping.")
}
```

Read in gtex RDS file 
```{r}
gtex=readRDS(here("data/bulk-gex_v8_rna-seq_GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.rds"))
```

Aggregate in-place by 'description'
```{r}
gtex <- as.data.table(gtex)
aggregated_gtex <- gtex[, lapply(.SD, sum), by = Description, .SDcols = is.numeric]
```

Create a meta data file to attach to the FBM matrix

```{r}
# For testing we will use 2000 columns
# set.seed(123);iisample=sample(2:ncol(aggregated_gtex),2000)
# meta=list()
# meta$rowNames=aggregated_gtex$Description
# meta$colNames=colnames(aggregated_gtex)[-1][iisample]
# data_mat <- as.matrix(aggregated_gtex[, -1])[, iisample] 

meta <- list()
meta$rowNames <- aggregated_gtex$Description
meta$colNames <- colnames(aggregated_gtex)[-1]
data_mat <- as.matrix(aggregated_gtex[, -1])
```

Create the FBM
```{r}
fbm_file <- file.path(output_nb_path, "FBMgtex")
#if already created
#gtexFBM <- FBM(nrow = nrow(aggregated_gtex), ncol = ncol(aggregated_gtex) - 1, backingfile = fbm_file)
gtexFBM <- FBM(nrow = nrow(data_mat), ncol = ncol(data_mat), backingfile = fbm_file, create_bk = T)
```

Populate it with data
```{r}
# Populate in chunks (much faster for large matrices)
block_size <- 100  # Adjust based on your system memory
n_blocks <- ceiling(nrow(aggregated_gtex) / block_size)

for (i in 1:n_blocks) {
  start_row <- (i-1) * block_size + 1
  end_row <- min(i * block_size, nrow(data_mat))
  
  gtexFBM[start_row:end_row, ] <- as.matrix(data_mat[start_row:end_row, ])
}
```

```{r}
cleanFBM(gtexFBM)
```

```{r}
rowStats=computeRowStatsFBM(gtexFBM)
plot(rowStats$row_means, rowStats$row_variances, log="y")
```
Filter the data. Doing aggressive filtering to reduce the number of genes.
```{r}
filterResult=filterFBM(gtexFBM, rowStats,mean_cutoff = 1, var_cutoff = 0.1, backingfile =paste0(fbm_file, "_filtered")) 
gtexFBMfiltered=filterResult$fbm_filtered
meta$rowNames=meta$rowNames[filterResult$kept_rows]
rowStatsAll=rowStats
rowStats$row_means=rowStats$row_means[filterResult$kept_rows]
rowStats$row_variances=rowStats$row_variances[filterResult$kept_rows]
```

Do the z-scoring
```{r}
zscoreFBM(gtexFBMfiltered, rowStats = rowStats)
```

save the FBM object for quick loading
```{r}
gtexFBMfiltered$save()
saveRDS(meta, paste0(fbm_file, "_meta.RDS"))
```

Precompute the SVD
```{r}
gtexFBM.svd=big_randomSVD(gtexFBMfiltered, k = 200)
```

Run simpleDecomp as the base model 
```{r}
decomp_file <- file.path(output_nb_path, "gtexSimpleDecompResult.RDS")

if (!file.exists(decomp_file)) {
  gtex.sdres <- simpleDecomp(gtexFBMfiltered, k = 200, svdres = gtexFBM.svd)
  saveRDS(gtex.sdres, file = decomp_file)
} else {
  gtex.sdres <- readRDS(decomp_file)
}
```

Download some pathways
```{r}
gmtList=list()

gmtList[["Wiki"]]=getGMT("https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=WikiPathway_2021_Human")
gmtList[["MGI"]]=getGMT("https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=MGI_Mammalian_Phenotype_Level_4_2021")
gmtList[["GO"]]=getGMT("https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=GO_Biological_Process_2023")
gmtList[["Reactome"]]=getGMT("https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=Reactome_2022")
gmtList[["CellMarkers"]]=getGMT("https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=CellMarker_2024")
```

Create a pathMat object and save
```{r}
pathMat = gmtListToSparseMat(gmtList[c(5)])
#saveRDS(gmtList, "gmtList.RDS")
#saveRDS(pathMat, "pathMat.RDS")
```

Get a matched pathway set
The FBM version doesn't do any subsetting, everything has to match exactly
```{r}
matchedPaths=getMatchedPathwayMat(pathMat, meta$rowNames)
Chat=getChat(matchedPaths, method = "fast")
```

# Run PLIER2

Run PLIER2, setting `max.iter` low for demonstration
```{r}
time_taken2 <- system.time({
  gtex.plier2 = PLIERv2(
    gtexFBMfiltered,
    matchedPaths,
    sdres = gtex.sdres,
    Chat = Chat,
    doCrossval = TRUE,
    multiplier = 3,
    max.U.updates = 3,
    max.iter = 350
  )
})

print(time_taken2)
save(time_taken2, file = file.path(output_nb_path, "gtex_PLIER2_time_taken.rds"))

saveRDS(gtex.plier2, file = file.path(output_nb_path, "gtex_PLIER2.rds"))
```

# Run PLIER

Run PLIER with the same inputs than PLIER2

```{r}
library(PLIER)

time_taken <- system.time({
  gtex.plier = PLIER::PLIER(
    gtexFBMfiltered[], 
    as.matrix(matchedPaths), 
    svdres = gtexFBM.svd, 
    Chat = as.matrix(Chat), 
    doCrossval = TRUE, 
    max.iter = 350, 
    k = 200
  )
})

print(time_taken)
save(time_taken, file = file.path(output_nb_path, "gtex_PLIER_time_taken.rds"))

saveRDS(gtex.plier, file = file.path(output_nb_path, "gtex_PLIER.rds"))
```

