# recount2 preprocessing

Load libraries
```{r}
library(bigstatsr)
library(data.table)
library(dplyr)
library(rsvd)
library(glmnet)
library(Matrix)
library(knitr)
library(here)
library(biomaRt)
#library(rhdf5)
#library(hdf5r)
```

Output 

```{r}
output_nb_path <- here("output/vignettes/plier2_testing/recount2")
dir.create(output_nb_path, showWarnings = FALSE, recursive = TRUE)
```

Load the local files

```{r}
source(here("R/solvers.R"))
source(here("R/utilsNew.R"))
```

Download recount2 

TODO: fix code here, recount2 data need to be downloaded and unpacked from
here: https://figshare.com/articles/dataset/recount_rpkm_RData/5716033/4?file=10881866
and then unpack and get the file recount_data_prep_PLIER.RDS
and recount_rpkm.RDS
which was the input for PLIER

```{r download_archs4_if_needed, eval = TRUE, echo = TRUE, message = TRUE}
# url <- "https://s3.dev.maayanlab.cloud/archs4/files/human_gene_v2.5.h5"
# output_dir <- here::here("data", "archs4")
# output_file <- file.path(output_dir, "human_gene_v2.5.h5")
# 
# if (!file.exists(output_file)) {
#   if (!dir.exists(output_dir)) {
#     dir.create(output_dir, recursive = TRUE)
#   }
#   command <- sprintf("wget -c '%s' -O '%s'", url, output_file)
#   tryCatch({
#     system(command)
#     cat("File downloaded successfully to:", output_file, "\n")
#   }, error = function(e) {
#     cat("Error during download:", e$message, "\n")
#   })
# } else {
#   cat("File already exists. Skipping download.\n")
# }
```

Read preprocessed recount2 data

```{r}
data_mat <- readRDS("data/recount2/recount_data_prep_PLIER.RDS")

meta <- list()
meta$gene_symbols <- rownames(data_mat$rpkm.cm)
meta$samples <- colnames(data_mat$rpkm.cm)

rm(data_mat)
```

Read in recount2 file

```{r}
rpkm.df <- readRDS("data/recount2/recount_rpkm.RDS")

# Transform ensembl id to genesymbol
mart <- biomaRt::useDataset("hsapiens_gene_ensembl",
                            biomaRt::useMart("ensembl"))
genes <- unlist(lapply(strsplit(rpkm.df$ENSG, "[.]"), `[[`, 1))
rpkm.df$ensembl_gene_id <- unlist(lapply(strsplit(rpkm.df$ENSG, "[.]"),
                                         `[[`, 1))
gene.df <- biomaRt::getBM(filters = "ensembl_gene_id",
                          attributes = c("ensembl_gene_id", "hgnc_symbol"),
                          values = genes,
                          mart = mart)
# filter to remove genes without a gene symbol
gene.df <- gene.df %>% dplyr::filter(complete.cases(.))
# add gene symbols to expression df
rpkm.df <- dplyr::inner_join(gene.df, rpkm.df,
                             by = "ensembl_gene_id")
# set symbols as rownames (req'd for PLIER)
rownames(rpkm.df) <- make.names(rpkm.df$hgnc_symbol, unique = TRUE)
# remove gene identifier columns
rpkm.df <- rpkm.df %>% dplyr::select(-c(ensembl_gene_id:ENSG))

data_mat <- rpkm.df[meta$gene_symbols, meta$samples]

rm(rpkm.df)

data_mat <- as.matrix(as.data.table(data_mat))
```

Create the FBM

```{r}
n_genes <- length(meta$gene_symbols)
n_samples <- length(meta$samples)

fbm_file <- file.path(output_nb_path, "FBMrecount2")

recount2FBM <- FBM(
  nrow        = n_genes,
  ncol        = n_samples,
  backingfile = fbm_file,
  create_bk   = TRUE
)
```

Populate it with data

```{r}
# Populate in chunks (much faster for large matrices)
block_size <- 100  # Adjust based on your system memory
n_blocks <- ceiling(n_genes / block_size)

for (i in 1:n_blocks) {
  start_row <- (i-1) * block_size + 1
  end_row <- min(i * block_size, nrow(data_mat))

  recount2FBM[start_row:end_row, ] <- as.matrix(data_mat[start_row:end_row, ])
}
```

```{r}
cleanFBM(recount2FBM)
```

```{r}
rowStats=computeRowStatsFBM(recount2FBM)
plot(rowStats$row_means, rowStats$row_variances, log="y")
```

Filter the data. Doing aggressive filtering to reduce the number of genes.

```{r}
filterResult=filterFBM(recount2FBM, rowStats,mean_cutoff = 1, var_cutoff = 0.1, backingfile =paste0(fbm_file, "_filtered")) 
recount2FBMfiltered=filterResult$fbm_filtered
meta$gene_symbols=meta$gene_symbols[filterResult$kept_rows]
rowStatsAll=rowStats
rowStats$row_means=rowStats$row_means[filterResult$kept_rows]
rowStats$row_variances=rowStats$row_variances[filterResult$kept_rows]
```

Do the z-scoring
```{r}
zscoreFBM(recount2FBMfiltered, rowStats = rowStats)
```

Save the FBM object for quick loading

```{r}
recount2FBMfiltered$save()
saveRDS(meta, paste0(fbm_file, "_meta.RDS"))
```

Precompute the SVD

```{r}
k = 800
recount2FBM.svd=big_randomSVD(recount2FBMfiltered, k = k)
```

Run simpleDecomp as the base model 

```{r}
decomp_file <- file.path(output_nb_path, "recount2SimpleDecompResult.RDS")

if (!file.exists(decomp_file)) {
  recount2.sdres <- simpleDecomp(recount2FBMfiltered, k = k, svdres = recount2FBM.svd)
  saveRDS(recount2.sdres, file = decomp_file)
} else {
  recount2.sdres <- readRDS(decomp_file)
}
```

Download some pathways

```{r}
gmtList=list()

gmtList[["Wiki"]]=getGMT("https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=WikiPathway_2021_Human")
gmtList[["MGI"]]=getGMT("https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=MGI_Mammalian_Phenotype_Level_4_2021")
gmtList[["GO"]]=getGMT("https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=GO_Biological_Process_2023")
gmtList[["Reactome"]]=getGMT("https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=Reactome_2022")
gmtList[["CellMarkers"]]=getGMT("https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=CellMarker_2024")
```

Create a pathMat object and save
```{r}
pathMat = gmtListToSparseMat(gmtList[c(5)])
#saveRDS(gmtList, "gmtList.RDS")
#saveRDS(pathMat, "pathMat.RDS")
```

Get a matched pathway set
The FBM version doesn't do any subsetting, everything has to match exactly
```{r}
matchedPaths=getMatchedPathwayMat(pathMat, meta$gene_symbols)
Chat=getChat(matchedPaths, method = "fast")
```

# Run PLIER2

Run PLIER2, setting `max.iter` low for demonstration
```{r}
time_taken2 <- system.time({
  recount2.plier2 = PLIERv2(
    recount2FBMfiltered,
    matchedPaths,
    sdres = recount2.sdres,
    Chat = Chat,
    doCrossval = TRUE,
    multiplier = 3,
    max.U.updates = 3,
    max.iter = 350,
    k = k
  )
})

print(time_taken2)
save(time_taken2, file = file.path(output_nb_path, "recount2_PLIER2_time_taken.rds"))

saveRDS(recount2.plier2, file = file.path(output_nb_path, "recount2_PLIER2.rds"))
```

# Run PLIER

Run PLIER with the same inputs than PLIER2

```{r}
library(PLIER)

time_taken <- system.time({
  recount2.plier = PLIER::PLIER(
    recount2FBMfiltered[], 
    as.matrix(matchedPaths), 
    svdres = recount2FBM.svd, 
    Chat = as.matrix(Chat), 
    doCrossval = TRUE, 
    max.iter = 350, 
    k = k
  )
})

print(time_taken)
save(time_taken, file = file.path(output_nb_path, "recount2_PLIER_time_taken.rds"))

saveRDS(recount2.plier, file = file.path(output_nb_path, "recount2_PLIER.rds"))
```

